---
title: "Simulation of Deep Sequenced Mixtures of Streptococcus pneumoniae Strains"
output:
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: console
---

The following Bash commands accurately describe the execution process for each transmission inference method. However, they are not configured to run directly from this Rmarkdown document. Instead, these commands were executed separately on a server. The Python scripts for each tool can be found in the scripts directory of the TRACS package.

## Generate simulations

```
for rep in {1..10}
do
  for d in 5 50 500
  do
    prefix=sim_d${d}_rep${rep}
    python tracs-sim.py -t 5 --genomes ./ref_genomes/*.fasta --tran-genome ./ref_genomes/6259_6#1.fasta -d $d -b 10000 -n 2 -o $prefix --prefix $prefix
  done
done

```

## Run tools

### TRACS

TRACS was run using representative genomes from each major Global Pneumococcal Sequencing Cluster described in Gladstone et al., *EBioMedicine* (2019).

```
for f in sim_d5*[0-9]
do
  echo "prefix  read1 read2" > ${f}/input_data.csv
  echo "sampleA ./${f}/${f}_A1.fastq.gz ./${f}/${f}_A2.fastq.gz" >> ${f}/input_data.csv
  echo "sampleB ./${f}/${f}_B1.fastq.gz ./${f}/${f}_B2.fastq.gz" >> ${f}/input_data.csv
  tracs pipe -i ${f}/input_data.csv -o ${f}_filt_newDB -t 15 --database gpsDB.zip --keep-all --filter
done
```

### InStrain

inStrain v1.8.0 was run using a single *S. pneumoniae* reference genome (SRR9998163) to align to.  

```
for f in sim_d5*[0-9]
do
  python run-instrain.py -t 5 --inputA ${f}/${f}_A1.fastq.gz ${f}/${f}_A2.fastq.gz --inputB ${f}/${f}_B1.fastq.gz ${f}/${f}_B2.fastq.gz --references SRR9998163_WGS_of_strep_pneumoniae_Serotype_19F_assembly.fasta -o ${f}_instrain -p ${f}
done
```

### StrainPhlAn

MetaPhlAn & StrainPhlAn v4.0.5 were run using the default database (mpa_vJan21_CHOCOPhlAnSGB_202103). 

```
for f in sim_d5*[0-9]
do
  python run-strainphlan.py --inputA ${f}/${f}_A1.fastq.gz ${f}/${f}_A2.fastq.gz --inputB .${f}/${f}_B1.fastq.gz ${f}/${f}_B2.fastq.gz -o ${f}_default_strainphlan -p ${f} -t 1 --refDB ./metaphlanDB
done
```

### StrainGE

A StrainGE database was built using the same GPS reference genomes as TRACS following the user manual by running

```
for f in *.fna
do
  straingst kmerize -o ${f%.fna}.hdf5 $f
done

straingst kmersim --all-vs-all -t 4 -S jaccard -S subset *.hdf5 > similarities.tsv
straingst cluster -i similarities.tsv -d -C 0.99 -c 0.90 --clusters-out clusters.tsv *.hdf5 > references_to_keep.txt
straingst createdb -o pan-genome-db.hdf5 *.hdf5
```

StrainGE v1.3.3 was then run on each simulation

```
for f in sim_d5*[0-9]
do
  python run_strainge.py --inputA ${f}/${f}_A1.fastq.gz ${f}/${f}_A2.fastq.gz --inputB ${f}/${f}_B1.fastq.gz ${f}/${f}_B2.fastq.gz --referenceDB ./strainge_gpsDB/pan-genome-db.hdf5 --ref-dir ./strainge_gpsDB/ --ref-sim ./strainge_gpsDB/similarities.tsv -o ${f}/${f}
done
```

## Compare results and generate plot

```{r, warning=FALSE, message=FALSE}
library(data.table)
library(tidyverse)
library(ggthemes)
library(ggsci)
```

```{r}
strainphlan <- map_dfr(Sys.glob("./data/pneumo_sim_mixtures/*strainphlan_distances.tsv"), ~{
  name <- gsub(".*/sim_", "", gsub("_strainphlan_.*", "", .x))
  df <- fread(.x) %>% 
    as_tibble() %>%
    add_column(sample=name, .before = 1)
  df$sim_dist <- as.numeric(gsub(".*d", "", gsub("_rep.*", "", name)))
  return(df)
})

splhan_lengths <- fread("./data/pneumo_SGB_lengths.csv") %>% as_tibble()
splhan_lengths <- splhan_lengths[!duplicated(splhan_lengths$strain),]
splhan_lengths$length/2e6

strainphlan <- tibble(
  simulation = gsub("sim_", "", strainphlan$sample),
  # min_dist = strainphlan$snp_dist,
  min_dist = strainphlan$snp_dist*2e6/splhan_lengths$length,
  simdist = strainphlan$sim_dist,
  method = "strainphlan"
)

### Strainge

strainge <- map_dfr(Sys.glob("./data/pneumo_sim_mixtures/*_strainge_comparison.tsv"), ~{
  
  name <- gsub("_strainge_comparison.*", "", gsub(".*/sim_", "", .x))
  
  df <- fread(.x) %>% 
    as_tibble() %>%
    group_by(ref) %>%
    summarise(
      min_dist=sum(common-sharedAlleles)
    ) %>%
    add_column(simdist=as.numeric(gsub(".*d", "", gsub("_rep.*", "", name))), .before=2) %>%
    add_column(simulation=name, .before = 2) %>%
    add_column(method="Strainge")
    
  df <- df[which.min(df$min_dist),]
  
  return(df)
})

strainge$ref <- NULL

sim_props <- do.call(rbind, purrr::map(Sys.glob("./data/pneumo_sim_mixtures/sim_dist_props/*"), ~{
  name <- gsub("_dist_props.*", "", gsub(".*/sim_", "", .x))
  fread(.x) %>% 
    as_tibble() %>%
    add_column(simulation=name, .before=1)
}))

distances <- do.call(rbind, purrr::map(Sys.glob("./data/pneumo_sim_mixtures/*_tracm_distances.csv"), ~{
  name <- gsub("_tracm_distances.*", "", gsub(".*/sim_", "", .x))
  fread(.x) %>% as_tibble() %>%
    add_column(simulation=name, .before=1)
}))

min_distances <- distances %>% 
  group_by(simulation) %>%
  summarise(
    min_dist = min(`filtered SNP distance`)
  )
min_distances$simdist <- as.numeric(gsub("d","", gsub("_.*","",min_distances$simulation)))
min_distances$method <- "tracs"


instrain_distances <- do.call(rbind, purrr::map(Sys.glob("./data/pneumo_sim_mixtures/*instrain_distances.csv"), ~{
  name <- gsub("_instrain.*", "", gsub(".*/sim_", "", .x))
  fread(.x) %>% as_tibble() %>%
    add_column(simulation=name, .before=1)
})) %>% 
  group_by(simulation) %>%
  summarise(
    min_dist = min(`population_SNPs`)
  )
instrain_distances$simdist <- as.numeric(gsub("d","", gsub("_.*","",instrain_distances$simulation)))
instrain_distances$method <- "instrain"

all_dists <- rbind(min_distances[min_distances$simulation %in% instrain_distances$simulation,], 
                   strainphlan[strainphlan$simulation %in% instrain_distances$simulation,],
                   strainge[strainge$simulation %in% instrain_distances$simulation,],
                   instrain_distances)


tb <- table((sim_props %>% filter(propA>0 & propB>0))$simulation)

all_dists$multi_same_strain <- all_dists$simulation %in% names(tb)[tb>1]

# Fix names
all_dists$sim_name <- paste('Simulated\nSNPs:', all_dists$simdist)
all_dists$Method <- all_dists$method
all_dists$Method[all_dists$method=='instrain'] <- 'inStrain'
all_dists$Method[all_dists$method=='tracs'] <- "TRACS"
all_dists$Method[all_dists$method=='Strainge'] <- 'StrainGE'
all_dists$Method[all_dists$method=='strainphlan'] <- 'StrainPhlAn'

# Calculate errors
error_pdf <- all_dists
error_pdf$error <- error_pdf$min_dist - error_pdf$simdist
error_pdf$rel_error <- (error_pdf$min_dist - error_pdf$simdist)/error_pdf$simdist

# Generate plot
dodge_width <- 0.7
gg <- ggplot(error_pdf, aes(x=simdist, y=rel_error, color=Method, group=Method)) + 
  scale_y_continuous(trans = scales::pseudo_log_trans(sigma = 1, base = 10),
                     breaks = c(-500,-50, -5, 0, 5, 50, 500, 5000),
                     limits = c(-5, 5000)) +
  scale_x_continuous(trans = scales::pseudo_log_trans(sigma = 1, base = 10),
                     breaks=c(0, 5, 50, 500)) +
  geom_hline(yintercept = 0, colour='black', size=0.5) +
  geom_boxplot(aes(group=interaction(simdist, method)), 
               position=position_dodge(width=dodge_width), outlier.colour = NA) +
  geom_point(position=position_jitterdodge(jitter.width = 0.05, dodge.width = dodge_width)) +
  ggsci::scale_color_nejm() +
  xlab("Simulated SNP distance") +
  ylab("Relative error in SNP distance") +
  theme_clean(base_size = 25) +
  theme(plot.background = element_blank()) 

gg

# ggsave("./figures/pneumo_simulation_snpdist_2d_scaled_nomidas_simple.png", width = 10, height = 7)
# ggsave("./figures/pneumo_simulation_snpdist_2d_scaled_nomidas_simple.pdf", width = 10, height = 7)
```


```{r}
sim_props$min_prop <- pmin(sim_props$propA, sim_props$propB)
sim_props <- sim_props %>%
  filter(distance <=500)

error_pdf$min_prop <- sim_props$min_prop[match(error_pdf$simulation, sim_props$simulation)]

# Generate plot
dodge_width <- 0.7
error_pdf$simulation_dist <- paste0("Simulated transmission\ndistance: ", error_pdf$simdist)
gg <- ggplot(error_pdf, aes(x=min_prop, y=rel_error, color=Method, group=Method)) + 
  scale_y_continuous(trans = scales::pseudo_log_trans(sigma = 1, base = 10),
                     breaks = c(-500,-50, -5, 0, 5, 50, 500, 5000),
                     limits = c(-5, 5000)) +
  scale_x_continuous(limits = c(0,1),
                     breaks=c(0, 0.25, 0.5, 0.75, 1)) +
  geom_hline(yintercept = 0, colour='black', size=0.5) +
  geom_point() +
  facet_wrap(~simulation_dist, scales = "free", nrow=1) +
  ggsci::scale_color_nejm() +
  xlab("Minimum strain frequency in transmission pair") +
  ylab("Relative error in SNP distance") +
  theme_clean(base_size = 25) +
  theme(plot.background = element_blank()) 

gg


ggsave("./figures/pneumo_simulation_snpdist_vs_frequency.png", width = 12, height = 7)
ggsave("./figures/pneumo_simulation_snpdist_vs_frequency.pdf", width = 12, height = 7)
```


## Comparing parameters

```
rm tracs_cmds_revision.txt
rm parallel_tracs_nocovfilt_cmds.txt
for f in sim_d5*[0-9]
do
  echo tracs pipe -i ${f}/input_data.csv -o tracs_nocovfilt_${f} -t 5 --database /data1/gerryt/tracm-data/simulations/pneumo_art_mix/gpsDB.zip --keep-all --filter --keep-cov-outliers >> tracs_cmds_revision.txt
done

parallel --joblog parallel_tracs_nocovfilt_cmds.txt --progress -j 30 < tracs_cmds_revision.txt

for f in tracs_nocovfilt*
do
#remove tracs_nocovfilt_
prefix=$(echo $f | sed 's/tracs_nocovfilt_//')
echo $prefix
cp ${f}/transmission_distances.csv ./tracs_nocovfilt/${prefix}_tracs_nocovfilt_distances.csv
done

#tar gzip 
tar -czf tracs_nocovfilt.tar.gz tracs_nocovfilt
```

Generate plot with different parameter settings

```{r}
tracs_distances <- do.call(rbind, purrr::map(Sys.glob("./data/pneumo_sim_mixtures/*_tracm_distances.csv"), ~{
  name <- gsub("_tracm_distances.*", "", gsub(".*/sim_", "", .x))
  fread(.x) %>% as_tibble() %>%
    add_column(simulation=name, .before=1) %>% 
  group_by(simulation) %>%
  slice_min(`filtered SNP distance`, with_ties = FALSE)
}))

nocov_distances <- do.call(rbind, purrr::map(Sys.glob("./data/pneumo_sim_mixtures/tracs_nocovfilt/*_tracs_nocovfilt_distances.csv"), ~{
  name <- gsub("_tracs_nocovfilt_distances.*", "", gsub(".*/sim_", "", .x))
  fread(.x) %>% as_tibble() %>%
    add_column(simulation=name, .before=1) %>% 
  group_by(simulation) %>%
  slice_min(`filtered SNP distance`, with_ties = FALSE)
}))

diff_params_dists <- tibble(
  simulation = c(tracs_distances$simulation, tracs_distances$simulation, nocov_distances$simulation, nocov_distances$simulation),
  `parameter set` = factor(rep(c('coverage & filter', 'coverage & no filter', 'filter & no coverage', 'no coverage & no filter'), 
                        each = nrow(tracs_distances)),
                        levels=c('coverage & filter', 'coverage & no filter', 'filter & no coverage', 'no coverage & no filter')),
  distance = c(tracs_distances$`filtered SNP distance`, 
               tracs_distances$`SNP distance`,
               nocov_distances$`filtered SNP distance`,
               nocov_distances$`SNP distance`)
)

diff_params_dists$`simulation distance` <- as.numeric(gsub("d","", gsub("_.*","",diff_params_dists$simulation)))




diff_params_dists <- diff_params_dists %>%
  filter(simulation %in% instrain_distances$simulation)

# Calculate errors
diff_params_error_pdf <- diff_params_dists
diff_params_error_pdf$error <- diff_params_error_pdf$distance - diff_params_error_pdf$`simulation distance`
diff_params_error_pdf$rel_error <- (diff_params_error_pdf$distance - diff_params_error_pdf$`simulation distance`)/diff_params_error_pdf$`simulation distance`

# Generate plot
dodge_width <- 1
gg <- ggplot(diff_params_error_pdf |> filter(`simulation distance`==5), aes(x=`parameter set`, y=error, color=`parameter set`, group=`parameter set`)) + 
  # scale_y_continuous(trans = scales::pseudo_log_trans(sigma = 1, base = 10),
  #                    breaks = c(-500,-50, -5, 0, 5, 50, 500, 5000),
  #                    limits = c(-5, 5000)) +
  # scale_x_continuous(trans = scales::pseudo_log_trans(sigma = 1, base = 10),
  #                    breaks=c(0, 5, 50, 500)) +
  geom_hline(yintercept = 0, colour='black', size=0.5) +
  geom_boxplot(aes(group=interaction(`simulation distance`, `parameter set`)), 
               position=position_dodge(width=dodge_width), outlier.colour = NA) +
  geom_point(position=position_jitterdodge(jitter.width = 0.05, dodge.width = dodge_width)) +
  ggsci::scale_color_nejm() +
  xlab("Simulated SNP distance") +
  ylab("Error in SNP distance") +
  theme_clean(base_size = 25) +
  theme(plot.background = element_blank()) 

gg


ggsave("./figures/pneumo_simulation_snpdist_diff_params.png", width = 10, height = 7)
ggsave("./figures/pneumo_simulation_snpdist_diff_params.pdf", width = 10, height = 7)
```
