---
title: "Comparison of CPU and memory resource use of each algorithm"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

It is difficult to directly compare the resource requirements of each algorithm because each one involves different parameters and often requires selecting specific programs for tasks like read alignment or inferring SNP distances from multiple sequence alignments. To enhance comparability, we consistently opted for the same algorithms used in the TRACS pipeline whenever feasible.

To assess speed and memory requirements, we ran each algorithm on a representative example of metagenomic sequencing data involving a pair of samples, where one genome had been transmitted. Some algorithms, like StrainGE, do not scale as efficiently when analysing multiple species simultaneously, which means that this algorithm is likely to take longer with larger datasets involving the transmission of multiple species.

Overall, these results offer a general indication of each algorithm's performance. However, the exact differences in speed and memory usage will vary depending on the characteristics of each specific dataset.

## Run each method

CPU and memory requirements were recorded using the unix `time` command.

#### TRACS

```
/usr/bin/time -o ./time_tests/tracs_full_pair_time.txt -v tracs pipe -i /data1/gerryt/tracm-data/simulations/meta_revision/sim_100/sim_d500_ref_GCF_000020225.1_ASM2022v1_genomic/input_data.csv -o temp_tracs -t 1 --database /data1/gerryt/tracm-data/simulations/meta_revision/tracs_db_same_genome_as_sim.zip --keep-all --filter 
```

#### InStrain

Note: As the focus of this work is the strain tracking algorithms we have only included assembly to provide a rough indication of the additional overhead methods that rely on custom databases require. Different assembly algorithms will result in different resource requirements.

```
/usr/bin/time -o ./time_tests/instrain_full_pair_time.txt -v python /data1/gerryt/tracs/scripts/run-instrain.py -t 1 --inputA /data1/gerryt/tracm-data/simulations/meta_revision/sim_100/sim_d500_ref_GCF_000020225.1_ASM2022v1_genomic/sim_d500_ref_GCF_000020225.1_ASM2022v1_genomic_A1.fastq.gz /data1/gerryt/tracm-data/simulations/meta_revision/sim_100/sim_d500_ref_GCF_000020225.1_ASM2022v1_genomic/sim_d500_ref_GCF_000020225.1_ASM2022v1_genomic_A2.fastq.gz --inputB /data1/gerryt/tracm-data/simulations/meta_revision/sim_100/sim_d500_ref_GCF_000020225.1_ASM2022v1_genomic/sim_d500_ref_GCF_000020225.1_ASM2022v1_genomic_B1.fastq.gz /data1/gerryt/tracm-data/simulations/meta_revision/sim_100/sim_d500_ref_GCF_000020225.1_ASM2022v1_genomic/sim_d500_ref_GCF_000020225.1_ASM2022v1_genomic_B2.fastq.gz --references /data1/gerryt/tracm-data/simulations/meta_revision/gtdb_rep_references/*.fna -o temp_instrain -p sim_d500_ref_GCF_000020225.1_ASM2022v1_genomic

/usr/bin/time -o ./time_tests/metaspades_full_pair_time2.txt -v metaspades.py -t 50 -1 /data1/gerryt/tracm-data/simulations/meta_revision/sim_100/sim_d500_ref_GCF_000020225.1_ASM2022v1_genomic/sim_d500_ref_GCF_000020225.1_ASM2022v1_genomic_combined_1.fastq.gz -2 /data1/gerryt/tracm-data/simulations/meta_revision/sim_100/sim_d500_ref_GCF_000020225.1_ASM2022v1_genomic/sim_d500_ref_GCF_000020225.1_ASM2022v1_genomic_combined_2.fastq.gz -o temp_metaspades2
```

#### StrainPlAn

```
/usr/bin/time -o ./time_tests/strainphlan_full_pair_time.txt -v python /data1/gerryt/tracs/scripts/run-strainphlan.py --inputA /data1/gerryt/tracm-data/simulations/meta_revision/sim_100/sim_d500_ref_GCF_000020225.1_ASM2022v1_genomic/sim_d500_ref_GCF_000020225.1_ASM2022v1_genomic_A1.fastq.gz /data1/gerryt/tracm-data/simulations/meta_revision/sim_100/sim_d500_ref_GCF_000020225.1_ASM2022v1_genomic/sim_d500_ref_GCF_000020225.1_ASM2022v1_genomic_A2.fastq.gz --inputB /data1/gerryt/tracm-data/simulations/meta_revision/sim_100/sim_d500_ref_GCF_000020225.1_ASM2022v1_genomic/sim_d500_ref_GCF_000020225.1_ASM2022v1_genomic_B1.fastq.gz /data1/gerryt/tracm-data/simulations/meta_revision/sim_100/sim_d500_ref_GCF_000020225.1_ASM2022v1_genomic/sim_d500_ref_GCF_000020225.1_ASM2022v1_genomic_B2.fastq.gz -o temp_strainphlan -t 1 --refDB /data1/gerryt/tracm-data/simulations/metaphlan/metaphlanDB
```

#### StrainGE

```
/usr/bin/time -o ./time_tests/strainge_full_pair_time.txt -v python /data1/gerryt/tracs/scripts/run_strainge.py --inputA /data1/gerryt/tracm-data/simulations/meta_revision/sim_100/sim_d500_ref_GCF_000020225.1_ASM2022v1_genomic/sim_d500_ref_GCF_000020225.1_ASM2022v1_genomic_A1.fastq.gz /data1/gerryt/tracm-data/simulations/meta_revision/sim_100/sim_d500_ref_GCF_000020225.1_ASM2022v1_genomic/sim_d500_ref_GCF_000020225.1_ASM2022v1_genomic_A2.fastq.gz --inputB /data1/gerryt/tracm-data/simulations/meta_revision/sim_100/sim_d500_ref_GCF_000020225.1_ASM2022v1_genomic/sim_d500_ref_GCF_000020225.1_ASM2022v1_genomic_B1.fastq.gz /data1/gerryt/tracm-data/simulations/meta_revision/sim_100/sim_d500_ref_GCF_000020225.1_ASM2022v1_genomic/sim_d500_ref_GCF_000020225.1_ASM2022v1_genomic_B2.fastq.gz --referenceDB /data1/gerryt/tracm-data/simulations/meta_revision/strainge_db/pan-genome-db.hdf5 --ref-dir /data1/gerryt/tracm-data/simulations/meta_revision/strainge_db/ --ref-sim /data1/gerryt/tracm-data/simulations/meta_revision/strainge_db/similarities.tsv -o temp_strainge
```

### Run comparison steps only

Here we run just the SNP distance estimate steps for each algorithm, assuming that the initial sample specific processing steps have already been completed. 

#### InStrain

```
/usr/bin/time -v -o ./time_tests/instrain_compare_pair_time.txt inStrain compare -i ./temp_instrain/sampleA ./temp_instrain/sampleB -o ./temp_instrain/temp_compaire --min_cov 2 -p 1
```

#### StrainGE

```
/usr/bin/time -v -o ./time_tests/strainge_compare_pair_time.txt straingr compare -o ./temp_strainge/temp_compare ./temp_strainge/sampleA/sampleA.hdf5 ./temp_strainge/sampleB/sampleB.hdf5
```


#### TRACS

```
mkdir temp_tracs/temp_compare
/usr/bin/time -v -o ./time_tests/tracs_combine_pair_time.txt tracs combine -i temp_tracs/sampleA temp_tracs/sampleB -o temp_tracs/temp_compare
/usr/bin/time -v -o ./time_tests/tracs_compare_pair_time.txt tracs distance --msa temp_tracs/temp_compare/*.fasta.gz -o temp_tracs/temp_compare/temp_compare.tsv --filter
```

## Generate summary plots

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(ggthemes)
library(ggpattern)
library(patchwork)
library(dplyr)
```

```{r}
res <- map_dfr(Sys.glob("./data/time_tests/*.txt"), ~{
  lines <- str_trim(read_lines(.x))
  tibble(
    method = gsub(".*/", "", gsub("_time.*", "", .x)),
    memory = as.numeric(str_split(lines[grepl("Maximum resident set size", lines)], ": ")[[1]][[2]])/(1024**2),
    cpu = sum(as.numeric(str_split_fixed(lines[grepl("^[US].*time*", lines)], ": ", 2)[,2]))/60/60
  )
})

res$memory[res$method == "tracs_compare_pair"] <- max(res$memory[res$method == "tracs_compare_pair"], res$memory[res$method == "tracs_combine_pair"])
res$memory[res$method == "tracs_compare_pair"] <- res$cpu[res$method == "tracs_compare_pair"] + res$cpu[res$method == "tracs_combine_pair"]
res <- res[res$method != "tracs_combine_pair",]

res$memory[res$method == "metaspades_full_pair"] <- max(res$memory[res$method == "metaspades_full_pair"], res$memory[res$method == "instrain_full_pair"])

assembly_cpu <- res$cpu[res$method == "metaspades_full_pair"]
res$cpu[res$method == "metaspades_full_pair"] <- assembly_cpu + res$cpu[res$method == "instrain_full_pair"]


res$Method <- recode(res$method,
                     "instrain_compare_pair" = "InStrain",
                     "instrain_full_pair" = "InStrain",
                     "strainphlan_full_pair" = "StrainPhlAn",
                     "strainge_compare_pair" = "StrainGE",
                     "strainge_full_pair" = "StrainGE",
                     "tracs_full_pair" ="TRACS",
                     "tracs_compare_pair" ="TRACS",
                     "metaspades_full_pair" = "InStrain (custom DB)")



res$compare <- ifelse(grepl("compare", res$method), "Compare", "Full")

# widen 
pdf <- res %>%
  filter(compare=='Full')

pdf$Method <- factor(pdf$Method, levels=c("InStrain (custom DB)", "InStrain", "StrainGE", "StrainPhlAn", "TRACS"))

gg1 <- ggplot(pdf, aes(x=Method, y=cpu, fill=Method)) +
  geom_bar(stat="identity") +
  geom_rect_pattern(
    data = subset(pdf, Method == "InStrain (custom DB)"),
    aes(
      xmin = as.numeric(Method) - 0.45,
      xmax = as.numeric(Method) + 0.45,
      ymin = cpu - assembly_cpu,
      ymax = cpu
    ),
    pattern = "crosshatch",
    pattern_fill = "black",
    pattern_density = 0.02,
    pattern_spacing = 0.02
  ) +
  labs(x="Method", y="Time (hours)") +
  theme_clean(base_size = 15) +
  theme(plot.background = element_blank()) +
  theme(legend.position = "none") +
  scale_fill_manual(values=c("TRACS"="#20854EFF", "InStrain"="#BC3C29FF", "InStrain (custom DB)"="#E38D81", 
                                       "StrainPhlAn"="#E18727FF", "StrainGE"="#0072B5FF"))


gg1




pdf <- res %>%
  filter(compare=='Compare')

pdf <- rbind(pdf, res %>%
               filter(method=='instrain_full_pair') %>%
               mutate(Method='InStrain (custom DB)'))

pdf$Method <- factor(pdf$Method, levels=c("InStrain (custom DB)", "InStrain", "StrainGE", "StrainPhlAn", "TRACS"))

gg2 <- ggplot(pdf, aes(x=Method, y=cpu*60*60, fill=Method)) +
  geom_bar(stat="identity") +
  theme_clean(base_size = 15) +
  theme(plot.background = element_blank()) +
  theme(legend.position = "none") +
  scale_fill_manual(values=c("TRACS"="#20854EFF", "InStrain"="#BC3C29FF", "InStrain (custom DB)"="#E38D81", 
                                       "StrainPhlAn"="#E18727FF", "StrainGE"="#0072B5FF")) +
  scale_y_continuous(trans = scales::pseudo_log_trans(base = 10), breaks = c(10, 100, 1e3, 1e4, 1e5)) +
  labs(x="Method", y="Time (Seconds)", title="CPU to compare a preprocessed sample pair") 

gg2



# Projected times for larger datasets

compare <- res %>%
  filter(compare=='Compare')

compare <- rbind(compare, res %>%
               filter(method=='instrain_full_pair') %>%
               mutate(Method='InStrain (custom DB)'))

compare <- rbind(compare, res %>%
               filter(method=='tracs_compare_pair') %>%
               mutate(Method='StrainPhlAn'))

full <- res %>%
  filter(compare=='Full')

#remove assembly from calculation
full$cpu[full$Method=="InStrain (custom DB)"] <- full$cpu[full$Method=="InStrain"]

# Estimate the time taken for preprocessing step per sample
compare$base <- (full$cpu[match(compare$Method, full$Method)] - compare$cpu)/2

time_projection <- function(n, base, comparison_time){
  return(n * 2*base + choose(n, 2) * comparison_time)
}

projected_times <- map_dfr(c(10, 100, 1000), ~{
  tibble(
    method = compare$Method,
    projected_time = time_projection(.x, compare$base, compare$cpu),
    n_pairs = .x
  )
})


gg3 <- ggplot(projected_times, aes(x=n_pairs, y=projected_time, colour=method)) +
  geom_jitter(size=7, height = 0, width=0.01) +
  scale_y_log10(breaks = c(10, 100, 1e3, 1e4, 1e5, 1e6, 1e7)) +
  scale_x_log10() +
  labs(x="Number of Samples", y="Projected time (hours)", title="Projected time to compair metagenomic sample pairs") +
  theme_clean(base_size = 15) +
  theme(plot.background = element_blank()) +
  theme(legend.position = "none") +
  scale_colour_manual(values=c("TRACS"="#20854EFF", 
                               "InStrain"="#BC3C29FF", 
                               "InStrain (custom DB)"="#E38D81", 
                               "StrainPhlAn"="#E18727FF", 
                               "StrainGE"="#0072B5FF"))
gg3


pdf <- res %>%
  filter(compare=='Full')

pdf$Method <- factor(pdf$Method, levels=c("InStrain (custom DB)", "InStrain", "StrainGE", "StrainPhlAn", "TRACS"))


gg4 <- ggplot(pdf, aes(x=Method, y=memory, fill=Method)) +
  geom_bar(stat="identity") +
  labs(x="Method", y="Memory (Gb)") +
  theme_clean(base_size = 15) +
  theme(plot.background = element_blank()) +
  # theme(legend.position = "none") +
  scale_fill_manual(values=c("TRACS"="#20854EFF", "InStrain"="#BC3C29FF", "InStrain (custom DB)"="#E38D81", 
                                       "StrainPhlAn"="#E18727FF", "StrainGE"="#0072B5FF"))
gg4



gg1 + gg2 + gg3 + gg4 + patchwork::plot_layout(nrow = 2, ncol = 2)


ggsave("./figures/resources.png", width = 12, height = 10)
ggsave("./figures/resources.pdf", width = 12, height = 10)
```